Resolvable Ambiguity

### 2019-07-11 paper/paper.tex
Add explanations to informal part in 6.3
Pass over Introduction

% Motivate why unambiguous parsing is a good (and important) idea.
Ever since the early 60s, it is known to be undecidable to determine whether a context-free
grammar is ambiguous~\cite{cantorAmbiguityProblemBackus1962}. As a consequence,
a large number of restricted grammars have been developed to guarantee
that a language definition is unambiguous. This traditional view of
only allowing unambiguous grammars has been taken for granted as the
only truth: \emph{the} way of how the syntax of a language must be
defined ~\cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}.
%
However, recent work by
~\citet{palmkvistCreatingDomainSpecificLanguages2019} shows why
carefully designed ambiguous syntax definitions are in some cases
preferable compared to completely unambiguous
definitions. Specifically, instead of forcing language designers to
make arbitrary choice to disambiguate the language, a more natural
choice may be to postpone the decision to the programmer.

% Motivation II, compising DSLs
Another area---where ambiguous grammars naturally arises---is
domain-specific language development.  An important
objective when designing domain-specific languages is to construct
languages by composing and extending existing
languages.
Unfortunately, the composition of restricted
%grammars easily results in ambiguous grammars, which again are
%undecidable.
%
%constructing domain-specific programming languages there is often
%a desire to reuse work, to compose smaller language
%fragments.
%However, how to make such compositions well-behaved is not
%always obvious.
%In particular, when constructing new domain-specific languages from
%existing languages, by composing multiple syntax definitions,
grammars easily produces
an ambiguous grammar, even if all of the original grammars are
unambiguous. Approaches for solving this problem can be broadly split
into two categories: those that handle ambiguities on the
grammar-level (detection, prevention, etc.), and those that work on
particular examples of ambiguous programs. The former is well explored
in the form of heuristics for ambiguity detection
\cite{bastenAmbiguityDetectionProgramming2011,axelssonAnalyzingContextFreeGrammars2008,brabrandAnalyzingAmbiguityContextFree2007}
or restrictions on the composed grammars
\cite{kaminskiModularWellDefinednessAnalysis2013}, while the
latter has received relatively little attention.

@@

This paper focuses on the problems that arise if a grammar definition
is not guaranteed to be unambiguous. Concretely, if parsing a program
results in several parse trees, the grammar is obviously
ambiguous. The programmer then needs to disambiguate the program by,
for instance, inserting extra parenthesis. We say that a program has a
\emph{resolvable ambiguity} if the ambiguity can be solved by
modifying the program, such that each of the possible parse trees can
be selected using different modifications.
We can divide the problem
related to resolvable ambiguity into several different decision
problems. The \emph{dynamic resolvable ambiguity problem} is as follows: given a parse function for a language, and a specific program that is parsed into a set of parse trees, decided whether it possible to rewrite the program, such that each of the parse trees can be generated unambiguously. This problem can be generalized into the \emph{static resolvable ambiguity problem}: decide whether all possible parsable programs are resolvably ambiguous. Static resolvablity means that a language designer can be confident that the end-user (the program designer) always can resolve an ambiguity by modifying the program. By contrast, dynamic resolvability means that a language designed knows if a \emph{specific} program is resolvable. Hence, static resolvability gives formally verified guarantees on the syntax definition, whereas dynamic resolvability enables the language designer to test the resolvability of the syntax definition, program by program.

@@

In this paper, we solve the static resolvability problems for certain classes of languages, and demonstrate practically how dynamic resolvability may be used when designing language syntaxes. More specifically, we make the following contributions:

\begin{itemize}
\item We formally define the term \emph{resolvable ambiguity}, as well as giving precise formalizations of the static and dynamic \emph{resolvable ambiguity problems} (Section~\ref{sec:resolvable-definition}).
\item We formulate the \emph{static resolvability problem} of increasing difficulty, by introducing two different restricted language versions. We formalize a decidable algorithm that is sound and complete for the first version, and sound for the second (Section~\ref{sec:static}).
\item We define a decidable algorithm that solves the \emph{dynamic resolvability problem} for general context-free languages with well-balanced parentheses (Section~\ref{sec:dynamic}).
\item The dynamic resolvability methodology is implemented as a tool and a small DSL for defining syntax definitions. Two different cases studies are evaluated using this methodology and tool (Section~\ref{sec:evaluation}): (i) an implementation and evaluation of a large subset of OCaml's syntax, where we study the effect of implementing an underspecified language definition, and (ii) a domain-specific language for orchestrating parallel computations, where we investigate the effects of composing language fragments.
\end{itemize}

>>>>

% Motivate why unambiguous parsing is a good (and important) idea.
Ever since the early 60s, it has been known that determining whether a context-free
grammar is ambiguous is undecidable~\cite{cantorAmbiguityProblemBackus1962}. As a consequence,
a large number of restricted grammars have been developed to guarantee
that language definitions are unambiguous. This traditional view of
only allowing unambiguous grammars has been taken for granted as the
only truth: \emph{the} way of how the syntax of a language must be
defined ~\cite{sudkampLanguagesMachinesIntroduction1997,ahoCompilersPrinciplesTechniques2006,webberModernProgrammingLanguages2003,cooperEngineeringCompiler2011,ginsburgAmbiguityContextFree1966}.
%
However, recent work by
~\citet{palmkvistCreatingDomainSpecificLanguages2019} shows why
carefully designed ambiguous syntax definitions are in some cases
preferable compared to completely unambiguous
definitions. Specifically, instead of forcing language designers to
make more or less arbitrary choices to disambiguate the language, a more natural
solution may be to postpone the decision to the programmer.

% Motivation II, compising DSLs
Another area where ambiguous grammars naturally arise is
in domain-specific language development.  An important
objective when designing domain-specific languages is to be able construct
languages by composing and extending existing
languages.
Unfortunately, the composition of
%grammars easily results in ambiguous grammars, which again are
%undecidable.
%
%constructing domain-specific programming languages there is often
%a desire to reuse work, to compose smaller language
%fragments.
%However, how to make such compositions well-behaved is not
%always obvious.
%In particular, when constructing new domain-specific languages from
%existing languages, by composing multiple syntax definitions,
grammars easily produces
an ambiguous grammar, even if the original grammars are all
unambiguous on their own. Approaches for solving this problem can be broadly split
into two categories: those that handle ambiguities on the
grammar-level (detection, prevention, etc.), and those that work on
particular examples of ambiguous programs. The former is well explored
in the form of heuristics for ambiguity detection
\cite{bastenAmbiguityDetectionProgramming2011,axelssonAnalyzingContextFreeGrammars2008,brabrandAnalyzingAmbiguityContextFree2007}
or restrictions on the composed grammars
\cite{kaminskiModularWellDefinednessAnalysis2013}, while the
latter has received relatively little attention.

@@

This paper focuses on the problems that arise if a grammar definition
is not guaranteed to be unambiguous. Concretely, if parsing a program
could result in several correct parse trees, the grammar is obviously
ambiguous. The programmer then needs to disambiguate the program by,
for instance, inserting extra parentheses. We say that a program is
\emph{resolvably ambiguous} if the ambiguity can be resolved by
modifying the program, such that each of the possible parse trees can
be selected using different modifications.
%
We can divide the problem of resolvable ambiguity into two
different decision problems.
%
The \emph{dynamic resolvable ambiguity problem} is as follows:
given a parse function for a language, and a specific program that
is parsed into a set of parse trees, decide whether it possible to
rewrite the program, such that each parse tree can be generated
unambiguously.
%
This problem can be generalized into the \emph{static resolvable
  ambiguity problem}: given a parse function, decide whether all
parsable programs are resolvably ambiguous.
%
Static resolvablity means that a language designer can be
confident that the end-user (the programmer) can always resolve an
ambiguity by modifying the program. Identifying such ambiguities
for a \emph{specific} program and suggesting disambiguations is
enabled by dynamic resolvability.
%
Static resolvability gives formally verified guarantees on a
syntax definition, whereas dynamic resolvability enables the
language designer to test the resolvability of the syntax
definition, program by program.

@@

In this paper, we solve the static resolvability problems for certain classes of languages, and demonstrate practically how dynamic resolvability may be used when defining language syntax. More specifically, we make the following contributions:

\begin{itemize}
\item We formally define the term \emph{resolvable ambiguity}, as
  well as giving precise formalizations of the static and dynamic
  \emph{resolvable ambiguity problems}
  (Section~\ref{sec:resolvable-definition}).
\item We look at versions of the \emph{static resolvability
    problem} with increasing difficulty by considering two
  differently restricted classes of languages. We formalize a
  decidable algorithm that is sound and complete for the first
  version, and sound for the second (Section~\ref{sec:static}).
\item We define a decidable algorithm that solves the
  \emph{dynamic resolvability problem} for general context-free
  languages with well-balanced parentheses
  (Section~\ref{sec:dynamic}).
\item We evaluate the dynamic resolvability methodology in two
  different case studies using a tool and a small DSL for defining
  syntax definitions (Section~\ref{sec:evaluation}): (i) a
  domain-specific language for orchestrating parallel
  computations, where we investigate the effects of composing
  language fragments, and (ii) an implementation and evaluation of
  a large subset of OCaml's syntax, where we study the effect of
  implementing an underspecified language definition.
\end{itemize}
